---
title: "Time Series Final Project"
author: "Asad Adnan, Muhammad Ahmad, Brian Murphy"
date: "2025-02-17"
output: html_document
---


Crime Rate Forecasting

The objective of this project is to analyze and forecast crime rates using historical crime
data for Chicago. By applying time series forecasting methods, such as ARIMA, Seasonal ARIMA, and
regression models, the goal is to predict future crime trends to assist law enforcement and
policymakers in improving resource allocation and crime prevention strategies. Forecasting
crime rates will help Chicago better prepare for high-crime periods and allocate resources to
areas with the highest predicted need.


# Load necessary libraries
```{r}
library(forecast)
library(readxl)
library(ggplot2)
library(dplyr)
library(lubridate)
library(xgboost)
library(prophet)
```

# Load the dataset
```{r}
crime_data <- read.csv("C:/Users/asada/OneDrive/Desktop/Spring 1/TSF/Final Project/Crimes_-_2001_to_Present.csv")

head(crime_data)
```

# List of columns to drop
```{r}
columns_to_drop <- c('Block', 'Description', 'IUCR', 'Location.Description', 
                     'Beat', 'District', 'Ward', 'Community.Area', 'FBI.Code', 
                     'X.Coordinate', 'Y.Coordinate', 'Updated.On', 'Latitude', 
                     'Longitude', 'Location', 'Domestic', 'Year', 'Case.Number')
filtered_crime_data <- crime_data[ , !(names(crime_data) %in% columns_to_drop)]

head(filtered_crime_data)
```

# Top crime types
```{r}
filtered_crime_data$Primary.Type <- as.character(filtered_crime_data$Primary.Type)

crime_counts <- filtered_crime_data %>%
  group_by(Primary.Type) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

head(crime_counts)
```

# Clean and filter the data
```{r}
cleaned_data <- filtered_crime_data %>%
  filter(Primary.Type %in% c("THEFT", "BATTERY"), # Pick top two crime types
         Arrest == "true") %>% 
  arrange(Date) # Filter only 'THEFT' & 'BATTERY' with true Arrests & sort by the newly formatted Date

head(cleaned_data)
```

# Transform, and aggregate the data
```{r}
aggregated_data <- filtered_crime_data %>%
  mutate(Date = mdy_hms(Date)) %>%
  select(-Arrest, -ID) %>%   # Remove 'Arrest' and 'ID' columns
  mutate(Year = year(Date),  # Extract Year and Month for aggregation
         Month = month(Date)) %>%
  group_by(Year, Month) %>% # Group by Year and Month and count number of arrests
  summarise(Arrest_Count = n(), .groups = "drop") %>%
  mutate(Date = as.Date(paste(Year, Month, "01", sep = "-"))) %>%   # New Date column assuming 1st of months
  select(Date, Arrest_Count) %>%
  arrange(Date)

head(aggregated_data)
```

# Convert Arrest_Count into a time series object
```{r}
crime_ts <- ts(aggregated_data$Arrest_Count, 
               start = c(2001, 1), end = c(2024, 12),    
               frequency = 12)  

# Create and save the main time series plot
p1 <- autoplot(crime_ts) + 
  ggtitle("Monthly Arrest Counts Over Time") + 
  xlab("Year") + ylab("Number of Arrests") +
  theme_minimal()

# Display the plot
p1

# Save the plot
ggsave("crime_timeseries.png", plot = p1, width = 10, height = 6, dpi = 300)
```

## Forecast Combination

### Step 1: Data Partitioning: Training and Validation 
We are going to partition the data: train.ts from Jan 2001 to Dec 2019, and valid.ts is from Jan 2020 to Dec 2024.
```{r}
nValid <- 60 # Months in the validation set

# Split into training and validation sets
train.ts <- window(crime_ts, end = c(2019, 12))  
valid.ts <- window(crime_ts, start = c(2020, 1), end = c(2024, 12))   

# Create and save the training vs validation plot
p2 <- autoplot(train.ts) + 
  autolayer(valid.ts, series = "Validation", color = "red") +
  ggtitle("Training vs Validation Data") +
  xlab("Year") + 
  ylab("Number of Arrests") +
  theme_minimal()

# Display the plot
p2

# Save the plot
ggsave("train_validation_split.png", plot = p2, width = 10, height = 6, dpi = 300)
```

### Step 2: Train multiple forecasting models and forecast on validation data

#### Build a Regression model - Model 1
```{r}
# Train a regression model with trend, quadratic trend, and seasonality
crime.lm <- tslm(train.ts ~ trend + I(trend^2) + season)

# Forecast
crime.lm.forecast <- forecast(crime.lm, h = nValid, level = 0)

# Plot
autoplot(crime.lm.forecast) + 
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Quadratic Trend + Seasonality Regression Model Forecast") +
  xlab("Year") +
  ylab("Number of Arrests") +
  theme_minimal()
```

#### Build a ETS model - Model 2
```{r}
crime.ets <- ets(train.ts, model = "ZZZ")

crime.ets.forecast <- forecast(crime.ets, h = nValid, level = 0)

# Create and save the ETS model forecast plot
p3 <- autoplot(crime.ets.forecast) + 
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("ETS Model Forecast") +
  xlab("Year") +
  ylab("Number of Arrests") +
  theme_minimal()

# Display the plot  
p3

# Save the plot
ggsave("ets_forecast.png", plot = p3, width = 10, height = 6, dpi = 300)
```

#### Build an auto.arima() model - Model 3
```{r}
crime.arima <- auto.arima(train.ts)

crime.arima.forecast <- forecast(crime.arima, h = nValid, level = 0)

autoplot(crime.arima.forecast) + 
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Auto ARIMA Model Forecast") +
  xlab("Year") +
  ylab("Number of Arrests") +
  theme_minimal()
```

#### Build a NN model - Model 4
We will build a NN model with a few parameters
```{r}
# Define neural network parameters
p <- 11  # Number of previous time steps used for forecast
P <- 1   # Number of previous seasonal values to use
size <- 7 # Number of hidden nodes
repeats <- 20 # Number of iterations/epochs for training

crime.nnetar <- nnetar(train.ts, repeats = repeats, p = p, P = P, size = size)

crime.nnetar.forecast <- forecast(crime.nnetar, h = nValid)

autoplot(crime.nnetar.forecast) +
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Neural Network Model Forecast") +
  xlab("Year") +
  ylab("Number of Arrests") +
  theme_minimal()
```

#### Build a seasonal naive - Model 5
```{r}
crime.snaive <- snaive(train.ts, h = nValid, level = 0)

autoplot(crime.snaive) +
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Seasonal Naïve Model Forecast") +
  xlab("Year") +
  ylab("Number of Arrests") +
  theme_minimal()
```

### Step 3: Aggregate multiple forecasts

#### Simple Average
```{r}
# Define the number of models
num.models <- 5

# Compute the simple average of all model forecasts
crime.comb.simple.avg <- (crime.lm.forecast$mean + crime.ets.forecast$mean + 
                          crime.arima.forecast$mean + crime.nnetar.forecast$mean + 
                          crime.snaive$mean) / num.models

autoplot(train.ts) +
  autolayer(crime.comb.simple.avg, series = "Simple Avg Comb", color = "blue") +
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Simple Average of Forecast Models") +
  xlab("Year") +
  ylab("Number of Arrests") +
  theme_minimal()
```

#### Trimmed mean 
```{r}
# Collect the forecasts in a dataframe
forecast.vectors.df <- data.frame(cbind(crime.lm.forecast$mean, 
                                        crime.ets.forecast$mean, 
                                        crime.arima.forecast$mean, 
                                        crime.nnetar.forecast$mean, 
                                        crime.snaive$mean))

# Function to compute trimmed mean for each row separately - Using 20% trimming
# Since we have 5 models, we trim one highest and one lowest forecast
forecast.vectors.df$comb.trimmed.avg <- apply(forecast.vectors.df, 1, function(x) mean(x, trim = 0.2))

# Convert the object into a time series object
crime.comb.trimmed.avg <- ts(forecast.vectors.df$comb.trimmed.avg, start = c(2020, 1), end = c(2024, 12), freq = 12)

autoplot(train.ts) +
  autolayer(crime.comb.trimmed.avg, series = "Trimmed Avg Comb", color = "blue") +
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Trimmed Mean of Forecast Models") +
  xlab("Year") +
  ylab("Number of Crimes") +
  theme_minimal()
```

#### Running a regression that best fits the validation data

```{r}
# Collect the forecasts in a dataframe
forecast.vectors.df <- data.frame(cbind(crime.lm.forecast$mean, 
                                        crime.ets.forecast$mean, 
                                        crime.arima.forecast$mean, 
                                        crime.nnetar.forecast$mean, 
                                        crime.snaive$mean))

# Rename columns for clarity
colnames(forecast.vectors.df) <- c("lm_forecast", "ets_forecast", "arima_forecast", "nnetar_forecast", "snaive_forecast")

# Add validation data as another column
forecast.vectors.df$valid <- as.numeric(valid.ts)

# Running regression model to find the best combination
crime.forecasts.lm <- lm(valid ~ lm_forecast + ets_forecast + 
                           arima_forecast + nnetar_forecast + snaive_forecast, 
                         data = forecast.vectors.df)

# Summarizing the regression model
summary(crime.forecasts.lm)
```

#### Plotting the regression fit 

```{r}
# Convert the fitted values from regression model into a time series object
crime.comb.regression <- ts(crime.forecasts.lm$fitted.values, 
                            start = c(2020, 1), 
                            end = c(2024, 12), 
                            freq = 12)

# Create and save the regression combination model plot
p4 <- autoplot(train.ts) +
  autolayer(crime.comb.regression, series = "Regression Fit", color = "blue") +
  autolayer(valid.ts, series = "Observed", color = "red") +
  ggtitle("Crime Rate Forecasting - Best Regression Combination Model") +
  ylab("Number of Arrests") +
  xlab("Year") +
  theme_minimal()

# Display the plot
p4

# Save the plot
ggsave("regression_combination.png", plot = p4, width = 10, height = 6, dpi = 300)
```

### Step 4: Finally, compare the accuracy of all the models - MAPE
```{r}
# Compute MAPE for all models
c(
  LM = accuracy(crime.lm.forecast, valid.ts)["Test set", "MAPE"], 
  ETS = accuracy(crime.ets.forecast, valid.ts)["Test set", "MAPE"],
  ARIMA = accuracy(crime.arima.forecast, valid.ts)["Test set", "MAPE"],
  NNAR = accuracy(crime.nnetar.forecast, valid.ts)["Test set", "MAPE"],
  SNAIVE = accuracy(crime.snaive, valid.ts)["Test set", "MAPE"],
  comb.simple.avg = accuracy(crime.comb.simple.avg, valid.ts)["Test set", "MAPE"], 
  comb.reg = accuracy(crime.comb.regression, valid.ts)["Test set", "MAPE"]
)
```


